Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-08-01 Writing/manuscript.Rmd
finalized for second submission

---
title: Aquifer depletion exacerbates agricultural drought losses in the US High Plains
author:
  - Taro Mieno^[Department of Agricultural Economics, University of Nebraska Lincoln, tmieno2@unl.edu]
  - Timothy Foster^[Department of WWW, University of Manchester, timothy.foster@manchester.ac.uk]
  - Shunkei Kakimoto^[Department of Applied Economics, University of Minnesota, kakim002@umn.edu]
  - Nicholas Brozovic^[Department of Agricultural Economics, University of Nebraska Lincoln, nbrozovic@nebraska.edu]
abstract: |
  Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in drought years. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
acknowledgments: |
  Funding for the research in this manuscript was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007.
bibliography: DRA.bib
csl: iopart.csl
output:
  bookdown::pdf_document2:
    extra_dependencies: ["float"]
    citation_package: natbib
    keep_tex: true
    toc: false
    includes:
      in_header: "preamble.tex"
---

@@

# Main

Groundwater is an essential input for agricultural production, providing a critical buffer against limited or variable surface water supplies for farmers in many parts of the world [@scanlon2023global]. Globally, agriculture's dependence on groundwater for irrigation is expected to increase in the future because of higher crop water requirements, more erratic rainfall, and more frequent and extreme drought events caused by climate change [@zhou2010impact; @wada2013multimodel; @wada2014sustainability; @kreins2015quantification; @florke2018water]. However, many of the world's most important aquifer systems have experienced large reductions in storage over recent decades [@wada2010global; @famiglietti2011satellites; @scanlon2012groundwater; @konikow2015long; @bierkens2019non], which, if not addressed, will negatively affect the ability of farmers, food supply chains, and rural economies to use groundwater as a means of adapting to climate change.

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks [@schlenker2009nonlinear; @lobell2014greater; @schlenker2010robust; @zhou2020connections; @borgomeo2020impact]. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification to mitigate negative impacts of drought [@kuwayama2019estimating; @zipper2016drought; @zhu2022untangling; @zhu2022warming; @lu2020mapping; @davis2019sensitivity; @li2018changes]. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling [@foster2015well; @cotterman2018groundwater; @kahil2015modeling; @yoon2021coupled; @rad2020mod], with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage [@jain2021groundwater; @suter2021depletion].

Two pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability [@foster2015analysis]. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs [@mieno2017price; @bhattarai2021impact]. Second, depletion also reduces the pumping capacity of wells used to extract groundwater, which may create a physical constraint to the volume of irrigation that can be applied to a field [@konikow2005groundwater; @foster2014modeling; @hrozencik2017heterogeneous]. The combined effect of these two factors could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production [@foster2014modeling; @rad2020effects]. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change [@ukkola2020robust; @chiang2021evidence; @cook2020twenty]

In this paper, we present empirical evidence about how aquifer depletion affects farmers ability to effectively and reliably buffer crops against drought risk using data on historical crop yields, production areas, and aquifer conditions for the High Plains region in the United States. Our analysis extends previous research on the impacts of drought on agricultural production by evaluating how drought-related production losses are influenced by severity of groundwater scarcity, moving beyond prior binary comparisons of rainfed and irrigated production [@schlenker2009nonlinear; @lobell2014greater; @lu2018crop]. We also provide new empirical insights about the relative roles of extensive (i.e., irrigated area) and intensive (i.e., per-area irrigation rates) margin water use adjustments in determining drought-related production losses, which previously had only been demonstrated through theoretical modeling [@foster2014modeling; @foster2017effects; @rad2020effects]. Our findings provide important evidence of the value of preserving aquifer storage as a buffer against drought events. In doing so, our analysis adds weight to the need to improve groundwater resource conservation and sustainability to enable adaptation to climate change in groundwater-dependent agricultural systems globally [@jain2021groundwater; @scanlon2023global].

@@

## Aquifer depletion increases sensitivity of crop yields to drought {#impact-intensive}

@@

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \@ref(fig:yield-response)). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers' ability to fully meet crop water requirements during periods of more severe or extended precipitation deficits [@rad2020effects, @foster2014modeling; @hrozencik2017heterogeneous]. However, yield gradually increases as water deficit increases for counties with the highest level of aquifer thickness.

Figure \@ref(fig:yield-dif) shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95 percent confidence interval for corn an soybean. For corn, in years with water deficit higher than about 800 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference is `r round(test_results[balance == 1000 & crop == "Corn" & comp_q == 3, dif_hat], digits = 2)` tonne/ha. At low water deficits (below 200mm), counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \@ref(fig:yield-response)) experience statistically lower yields than counties with the lowest level of aquifer thickness (1st quantile). This counter-intuitive result may be due to the increased possibility of over-irrigation in regions with high aquifer thickness and well capacities [@foster2019assessing; @gibson2017case; @gibson2019benchmarking]. Unlike corn, soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding as corn and even the counties the lowest aquifer thickness level can catch up with water demand in severe drought years just as well as counties with higher aquifer thickness levels.

<!-- However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn. -->
```{r yield-response, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area irrigated yields of corn and soybean in US High Plains", out.width = "6.5in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_intensive.pdf"))
```

```{r yield-dif, out.width = "6in", fig.cap = "The difference in yield for the counties with the second and third (highest) aquifer thickness groups relative to the first (lowest) aquifer thickness group."}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_ir_yield_dif.pdf"))
```

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in drought conditions (more than 800mm water deficit), the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically [@rad2020effects; @foster2014modeling; @hrozencik2017heterogeneous; @deines2020transitions], but there has been little empirical research on farmers' irrigated area choices or how these are affected by combinations of drought exposure and groundwater scarcity.

Figure \@ref(fig:ir-share) shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean. Figure \@ref(fig:ir-share) demonstrates clearly that aquifer thickness has large and significant impacts on the share of irrigated production, in particular for corn. As aquifer thickness declines, the share of irrigated acres for corn decreases from around 0.8 (80\%) for an average county-level aquifer thickness of 150m to 0.5 (50\%) for an average county-level aquifer thickness of 12m. Reductions in the share of production area under irrigation is also evident for soybean, although the effect is less pronounced than for corn potentially due to soybean's lower water requirements and susceptibility to drought than corn [@zipper2016drought; @ruess2022irrigation].

```{r ir-share, fig.cap = "The impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean in the US High Plains (Note: shaded area represents 95 percent confidence interval)", out.width = "5in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share.pdf"))
```

@@

res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[(abs(balance - 950)) < 1, by = crop]
```
The overall effects of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where weights are production area shares) for corn and soybean at thee values of aquifer thickness: 12.2, 59.1, 106.1, and 153.0  meters^[Confidence intervals were not presented to avoid over-crowding the figures. Please see Appendix \@ref(reg-conf) for estimated average yield response curves with 95 percent confidence interval.]. While you can see similar trends as presented in Figure \@ref(fig:yield-response), the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced mainly dues to lower irrigated area for lower aquifer thickness level (rainfed production has a larger share).

Figure \@ref(fig:dif-tot-impact) shows the difference in average per-area production of the 59.1m, 106.1m, and 153.0m aquifer thickness levels relative to 12.2m with the 95 percent confidence interval for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when precipitation deficits are small (less than around 200mm and 300mm for 59.1m and 106.1m aquifer thickness, respectively). However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (i.e. greater than around 800mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average corn and soybean productivity are about 25\% and 15\% lower, respectively, when differentiating between areas with the highest (3rd quantile) and lowest (1st quantile) aquifer thickness. Further, notice that the decline in average per-area production at a seasonal water deficit of 950mm is not linear. Going from 153.0m aquifer thickness to 106.1m (decline of 46.9m) would result in `r round(res_at_950[sat == 502 & crop == "corn", avg_yield] - res_at_950[sat == 348 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to aquifer thickness when the aquifer thickness level is high.
Going from 106.1m aquifer thickness to 59.1m (decline of 47m) would result in `r round(res_at_950[sat == 348 & crop == "corn", avg_yield] - res_at_950[sat == 194 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. However, going from 59.1m aquifer thickness to 12.2m (decline of 46.9m) would result in `r round(res_at_950[sat == 194 & crop == "corn", avg_yield] - res_at_950[sat == 40 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. That is, the same degree of decline is much more damaging when the decline is from a low aquifer thickness level than a higher aquifer thickness level. This is primarily due to the non-linear impact of aquifer thickness on the share of irrigated corn acres as shown in Figure \@ref(fig:ir-share). When aquifer thickness level is high enough (e.g., 100m), a small decline does not reduce the share of irrigated acres. However, when aquifer thickness level is low (e.g., 40m), a small decline does reduce the share of irrigated acres.

```{r tot-impact, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_total_impact.pdf"))
```

```{r dif-tot-impact, fig.cap = "Difference in average yield relative to the 1st aquifer thickness quantile for corn and soybean", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_avg_yield_dif.pdf"))
```

@@

We demonstrate the important role of aquifer thickness in determining farmers ability to successfully buffer crops against drought risks in groundwater irrigated agricultural systems such as the High Plains of the United States. Previous research focused on impacts of drought on agricultural systems has largely focused either on risks for rainfed production [@schlenker2009nonlinear; @lobell2014greater; @schlenker2010robust; @zhou2020connections; @borgomeo2020impact] or for irrigated lands where water supply constraints are assumed to be minimal [@kuwayama2019estimating; @zipper2016drought; @zhu2022untangling; @zhu2022warming; @lu2020mapping; @davis2019sensitivity; @li2018changes; @luan2021combined]. Our analysis adds new empirical insights to this literature which show explicitly that reductions in aquifer thickness increase the vulnerability of irrigated agricultural production areas to drought events. Critically, this suggests that commonly adopted binary definitions of agricultural land areas into rainfed and irrigated categories are likely to underestimate exposure of crop production to drought risks, in particular in areas where water supplies for irrigation are insufficient to fully meet crop water needs in all years due to constraints imposed by hydrology, policies, and/or economics.

@@

Our findings also provide useful guidance for water managers and policymakers seeking to address unsustainable groundwater use in the US High Plains and globally. Consistent with prior theoretical modeling, we demonstrate that the relationship between aquifer thickness and drought impacts is highly non-linear when accounting for farmers decisions about irrigated production areas. This indicates that sustainable groundwater conservation efforts [@macewan2017hydroecon; @butler2020charting; @elshall2020groundwater] should be targeted in space and time to avoid depletion exceeding critical thresholds which impair farmers' ability to effectively buffer production against drought. Analysis such as ours can provide valuable evidence about these critical thresholds and tipping points to guide proactive management of groundwater stocks, and, in doing so, limit economically damaging transitions from irrigated to rainfed production [@foster2017effects; @deines2020transitions]. For example, several groundwater management agencies in the Texas portion of the HPA have in recent years set out targets to ensure  50\% of current aquifer storage remains in 50 years time (so called 50/50 rule) [@closas2018chronicle]. The effectiveness of such targets could likely be improved substantially by setting conservation targets based on empirical evidence about the level of aquifer thickness that is required to ensure adequate levels of drought protection, both now and with future climate change.

>>>>

---
title: Aquifer depletion exacerbates agricultural drought losses in the US High Plains
author:
  - Taro Mieno^[Department of Agricultural Economics, University of Nebraska Lincoln, tmieno2@unl.edu]
  - Timothy Foster^[Department of WWW, University of Manchester, timothy.foster@manchester.ac.uk]
  - Shunkei Kakimoto^[Department of Applied Economics, University of Minnesota, kakim002@umn.edu]
  - Nicholas Brozovic^[Department of Agricultural Economics, University of Nebraska Lincoln, nbrozovic@nebraska.edu]
abstract: |
  Aquifer depletion poses a major threat to the ability of farmers, food supply chains, and rural economies globally to use groundwater as a means of adapting to climate variability and change. Empirical research has demonstrated the large differences in drought risk exposure that exist between rainfed and irrigated croplands, but previous work commonly assumes water supply for the latter is unconstrained. In this paper, we evaluate how aquifer depletion affects the resilience of irrigated crop production to drought risk using over 30 years of data on historical corn and soybean yields, production areas, and aquifer conditions for the High Plains region in the United States. We show that aquifer depletion reduces the ability of farmers to sustain irrigated crop yields and production areas in years and locations with large growing season water deficits. Our findings demonstrate that drought-related production losses on irrigated croplands increase non-linearly with aquifer depletion, highlighting the need for proactive aquifer conservation interventions to support adaptation and resilience to future increases in rainfall variability under climate change.
acknowledgments: |
  Funding for the research in this manuscript was provided by the United States Department of Agriculture under contract numbers OCE 58-0111-20-007 and OCE 58-0111-21-007.

bibliography: DRA.bib
csl: iopart.csl
output:
  bookdown::pdf_document2:
    extra_dependencies: ["float"]
    citation_package: natbib
    keep_tex: true
    toc: false
    includes:
      in_header: "preamble.tex"
---

@@

# Main
Groundwater is an essential input for agricultural production, providing a critical buffer against limited or variable surface water supplies for farmers in many parts of the world [@scanlon2023global]. Globally, agriculture's dependence on groundwater for irrigation is expected to increase in the future because of higher crop water requirements, more erratic rainfall, and more frequent and extreme drought events caused by climate change [@zhou2010impact; @wada2013multimodel; @wada2014sustainability; @kreins2015quantification; @florke2018water]. However, many of the world's most important aquifer systems have experienced large reductions in storage over recent decades [@wada2010global; @famiglietti2011satellites; @scanlon2012groundwater; @konikow2015long; @bierkens2019non]. If not addressed, will negatively affect the ability of farmers, food supply chains, and rural economies to use groundwater as a means of adapting to climate change.

Despite widespread alarm about the risks that aquifer depletion poses to crop productivity and food security, empirical evidence is limited about how reductions in groundwater availability will alter farmers capacity to adapt successfully to drought and rainfall variability. Most empirical studies that have assessed impacts of drought on agricultural productivity have focused primarily on rainfed production areas due to the greater exposure of rainfed agriculture to drought-related shocks [@schlenker2009nonlinear; @lobell2014greater; @schlenker2010robust; @zhou2020connections; @borgomeo2020impact]. Existing studies also demonstrate the importance of existing irrigated areas or future irrigation intensification and expansion to mitigate negative impacts of drought [@kuwayama2019estimating; @zipper2016drought; @zhu2022untangling; @zhu2022warming; @lu2020mapping; @davis2019sensitivity; @li2018changes]. In contrast, most research assessing impacts of aquifer depletion on resilience of irrigated farmland to drought has relied on simulation modeling [@foster2015well; @cotterman2018groundwater; @kahil2015modeling; @yoon2021coupled; @rad2020mod], with more limited work that attempts to empirically evaluate change in drought-related production losses and risks on irrigated lands as a function of changing aquifer storage [@jain2021groundwater; @suter2021depletion].

Two pathways exist by which aquifer depletion will impact farmers' ability to access and use groundwater as a buffer against drought and rainfall variability [@foster2015analysis]. First, depletion increases the energy requirements and costs of pumping water, potentially making it less economically viable for farmers to satisfy crop water needs [@mieno2017price; @bhattarai2021impact]. Second, depletion also reduces the aquifer thickness of an aquifer, lowering the aquifer's transmissivity (i.e., the potential rate of water supply to wells). This latter effect reduces the rate at which water can be pumped and applied to a field as irrigation [@konikow2005groundwater; @foster2014modeling; @hrozencik2017heterogeneous]. If the reduction in well yields is substantial - for example as would occur when a large proportion of aquifer thickness is depleted [@hecox2002calculation; @korus2020depletion] - then a farmer may be unable to extract sufficient water to meet crop water demands, in particular during periods of limited rainfall and in the peak of the growing season when crops are most sensitive to water deficits [@foster2015well;@rouhi2020downside]. The combined effect of declining well yields and increasing pumping costs could force farmers to reduce irrigation depths, resulting in greater risk of drought-related crop stress. Alternatively, or in addition, a farmer may be forced to limit irrigated area to ensure crop water demands can be met, leading to reductions in total crop production [@foster2014modeling; @rad2020effects]. Understanding how these two responses vary as a function of aquifer conditions is crucial to guide groundwater management planning decisions, in particular given projected increases in frequency and intensity of droughts under climate change [@ukkola2020robust; @chiang2021evidence; @cook2020twenty].

In this paper, we present empirical evidence about how aquifer depletion affects farmers ability to effectively and reliably buffer crops against drought using data on historical crop yields, production areas, and aquifer conditions for the High Plains region in the United States. Our analysis extends previous research on the impacts of drought on agricultural production by evaluating how drought-related production losses are influenced by severity of groundwater scarcity, moving beyond prior binary comparisons of rainfed and irrigated production [@schlenker2009nonlinear; @lobell2014greater; @lu2018crop]. We also provide new empirical insights about the relative roles of extensive (i.e., irrigated area) and intensive (i.e., per-area irrigation rates) margin water use adjustments in determining drought-related production losses, which previously had only been demonstrated through theoretical modeling [@foster2014modeling; @foster2017effects; @rad2020effects]. Our findings provide important evidence of the value of preserving aquifer storage as a buffer against drought events. In doing so, our analysis adds weight to the need to improve groundwater resource conservation and sustainability to enable adaptation to climate change in groundwater-dependent agricultural systems globally [@jain2021groundwater; @scanlon2023global].

@@

## Aquifer depletion increases sensitivity of crop yields to water deficits {#impact-intensive}

@@

We now focus on comparing yield response to water deficit among the three aquifer thickness categories (all irrigated). First, for water deficits above around 400mm, yield changes become negative for counties with the lowest levels of aquifer thickness (i.e., 1st quantile in Figure \@ref(fig:yield-response)). This likely reflects the restrictions that lower aquifer thickness and associated reductions in well yields place on farmers ability to fully meet crop water requirements during periods of more severe or extended water deficits [@rad2020effects, @foster2014modeling; @hrozencik2017heterogeneous]. In contrast, yield continues to increase gradually as water deficit increases beyond 400 mm for counties with the highest level of aquifer thickness.

Figure \@ref(fig:yield-dif) shows the difference in yield of the second and third (highest) aquifer quantiles relative to the first (lowest) quantile with the 95% confidence interval for corn and soybean. For corn, in years with water deficit greater than about 700 mm, the highest aquifer thickness group has statistically significantly higher yields than the lowest aquifer thickness group. At an extreme water deficit level of 950 mm (average water deficit in 2012, the most extreme drought in our historical record), the difference in yields between the smallest and largest aquifer thickness groups is `r round(test_results[balance == 1000 & crop == "Corn" & comp_q == 3, dif_hat], digits = 2)` tonne/ha. Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower saturated thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions.

At low water deficits (below 200mm), Figure \@ref(fig:yield-dif) illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure \@ref(fig:yield-response) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils [@li2019excessive]. Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years [@foster2019assessing; @gibson2017case; @gibson2019benchmarking], exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation.

<!-- However, it is important to highlight that impacts of aquifer thickness on yields for negative deficits is much less pronounced than that for extreme droughts (i.e. high deficits), in particular for corn. -->
```{r yield-response, fig.cap = "The impact of water deficit and aquifer thickness on rainfed and irrigated per-area yields of corn and soybean in US High Plains. Irrigated yield responses to water deficits are dissagregated by aquifer thickness level, considering three quantile groups that capture the range of aquifer thickness values in our sample: (i) 9 m to 30.1 m; (ii) 30.1 m to 67.2 m; and (iii) 67.2 m to 215.8 m. The histograms on the top of each panel show the distribution of county-year water deficit values included in regressions for each crop.", out.width = "6.5in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_yield_intensive.pdf"))
```

```{r yield-dif, out.width = "6in", fig.cap = "The difference in estimated county-level irrigated corn (top panels) and soybean (bottom panels) yields for counties in the second and third (largest) aquifer thickness quantiles relative to the first (smallest) aquifer thickness quantile."}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_ir_yield_dif.pdf"))
```

@@

While aquifer thickness has a statistically significant impact on per-area irrigated crop yields in drought conditions (water deficit greter than 700mm), the magnitude of the yield impact is small when compared to differences between irrigated and rainfed production for the same weather conditions. One explanation for this result is that farmers may choose to retain a smaller proportion of land in irrigated production, so that limited groundwater pumping capacity can be used to adequately buffer crops against drought on land that remains under irrigation. Reductions in irrigated areas as an aquifer is depleted have been demonstrated theoretically [@rad2020effects; @foster2014modeling; @hrozencik2017heterogeneous; @deines2020transitions], but there has been little empirical research on farmers' irrigated area choices or how these are affected by combinations and magnitudes of weather risks and groundwater scarcity.

corn_share <- all_results[crop == "corn", share_pred_data][[1]]

share_high <-
  corn_share[which.min(abs(sat - 130)), ir_share_hat] %>%
  round(digits = 2)

share_low <-
  corn_share[which.min(abs(sat - 10)), ir_share_hat] %>%
  round(digits = 2)
```

Figure \@ref(fig:ir-share) shows the estimated relationship between aquifer thickness and the share of production area that is irrigated for corn and soybean after accounting for spatio-temporal variations in weather conditions and drought risk exposure across the aquifer. Figure \@ref(fig:ir-share) demonstrates that aquifer thickness has meaningfully large impacts on the share of irrigated production for both corn and soybean. As aquifer thickness declines, the share of irrigated acres for corn decreases from around `r share_high` (`r share_high * 100`%) for an average county-level aquifer thickness of 130m to `r share_low` (`r share_low * 100`%)  for an average county-level aquifer thickness of 10m (See Appendix \@ref(test-dif-share) for formal statistical testings). Reductions in the share of production area under irrigation is also evident for soybean, although the effect is less pronounced than for corn, potentially due to soybean's lower water requirements and susceptibility to drought relative to corn [@zipper2016drought; @ruess2022irrigation]

```{r ir-share, fig.cap = "The estimated impact of aquifer thickness on the share of agricultural production under irrigation for corn and soybean producing counties in the US High Plains (Note: shaded area represents 95 percent confidence interval)", out.width = "5in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_share.pdf"))
```

@@

res_at_950 <-
  all_results[, .(crop, avg_yield_pred_data)] %>%
  unnest(cols = c(avg_yield_pred_data)) %>%
  data.table() %>%
  .[crop == "corn", ] %>%
  .[(abs(balance - 950)) < 5, ]
```

The overall influence of aquifer depletion on drought-related production losses is a function of both per-area irrigated yield and irrigated area changes highlighted in Sections \@ref(impact-intensive) and \@ref(impact-share). Combining both responses, Figure \@ref(fig:tot-impact) shows the relationship between water deficit and average per-area production (weighted-average of irrigated and rainfed yields, where the weights are production area shares) for corn and soybean at four values of aquifer thickness: 10, 50, 90, and 130 meters. While similar trends are observed as in Figure \@ref(fig:yield-response), the difference in production loss caused by high water deficits among the aquifer thickness levels are much more pronounced due to lower estimated irrigated area share for smaller aquifer thickness values. For example, for an aquifer thickness of 10 m, reductions in average yields occur for smaller water deficits than are required to trigger reductions in per-area irrigated yields alone. Similarly, the overall magnitude of yield reduction is greater for average yields than for per-area irrigated yields. This is because low aquifer thickness values have a greater estimated share of crop area under rainfed production, which magnifies average yield declines with increasing water deficits.

Figure \@ref(fig:dif-tot-impact) shows the difference in average per-area productivity for 50 m, 90 m, and 130 m aquifer thickness levels relative to an aquifer thickness of 10 m for corn and soybean. For corn, aquifer thickness does not have a significant impact on productivity when water deficits are below about 350mm. However, beyond this threshold, higher levels of water deficit result in noticeable and statistically significant reductions to average corn production. This reflects the combination of lower irrigated yields under drought conditions, and, more importantly, reduced overall productivity due to smaller share of land under irrigation in areas and years where aquifer thickness is lower. Similar effects are also observed for soybean, although reductions in average production only occur at a higher deficit threshold (greater than around 600mm) and are smaller in overall magnitude. For example, for a seasonal water deficit of 950mm (average water deficit in 2012, the most extreme drought in our historical record), average per-area corn and soybean productivity are about 25% and 15% lower, respectively, when differentiating between areas with the 130m and 10m aquifer thicknesses.

Figure \@ref(fig:tot-impact) and Figure \@ref(fig:dif-tot-impact) also demonstrate that impact of aquifer thickness on average per-area production is highly non-linear at high seasonal water deficits. At a water deficit of 950mm, for example, reducing from 130m aquifer thickness to 90m (decline of 40m) would result in only `r round(res_at_950[sat == 130 & crop == "corn", avg_yield] - res_at_950[sat == 90 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. This negligible difference in average yield from the decline is due to the insensitivity of the share of irrigated acres to a decline in aquifer thickness when the initial aquifer thickness level is high as can be seen in Figure \@ref(fig:ir-share). Reducing from 90m aquifer thickness to 50m (decline of 40m) would result in `r round(res_at_950[sat == 90 & crop == "corn", avg_yield] - res_at_950[sat == 50 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn, with a further drop from 50m aquifer thickness to 10m (decline of 40m) leading in `r round(res_at_950[sat == 50 & crop == "corn", avg_yield] - res_at_950[sat == 10 & crop == "corn", avg_yield], digits = 2)` tonne/ha of production loss for corn. Similar but less pronounced effects are also observed for soybean. This non-linear impact of aquifer thickness is due to the influence of aquifer thickness on the share of irrigated corn acres (Figure \@ref(fig:ir-share)), which implicitly reflects how aquifer thickness alters well yields and, in turn, farmers ability to effectively and reliably buffer crops against water deficits during the growing season.

```{r tot-impact, fig.cap = "Average productivity of corn and soybean in the US High Plains for different levels of water deficit and aquifer thickness, considering both per-area yield responses and farmers choices about irrigated area shares in response to different water deficit and aquifer conditions. See Appendix A for estimated average yield response curves with 95 percent confidence intervals displayed, which are omitted here for visual clarity.", out.width = "6in"}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_total_impact.pdf"))
```

```{r dif-tot-impact, fig.cap = "Difference in average yield of corn and soybean for aquifer thickness levels of 50 m, 90 m, and 130 m, relative to a baseline aquifer thickness of 10 m. Shaded areas denote 95 percent confidence interval, which when greater than zero indicate average yields for the given aquifer thickness level are statistically significantly greater than for an aquifer thickness of 10 m at the 5 percent level of significance.", out.width = "6in", cache = FALSE}
knitr::include_graphics(here::here("GitControlled/Writing/Figures/g_avg_yield_dif.pdf"))
```

@@

We demonstrate the important role of aquifer thickness in determining farmers ability to successfully buffer crops against drought risks in groundwater irrigated agricultural systems such as the High Plains of the United States. Previous research focused on impacts of drought on agricultural systems has largely focused either on risks for rainfed production [@schlenker2009nonlinear; @lobell2014greater; @schlenker2010robust; @zhou2020connections; @borgomeo2020impact] or for irrigated lands where water supply constraints are assumed to be minimal [@kuwayama2019estimating; @zipper2016drought; @zhu2022untangling; @zhu2022warming; @lu2020mapping; @davis2019sensitivity; @li2018changes; @luan2021combined]. Our analysis adds new empirical insights to this literature by quantifying how reductions in aquifer thickness increase the vulnerability of irrigated agricultural production areas to drought events. Critically, our findings suggest that commonly adopted binary definitions of agricultural land areas into rainfed and irrigated categories are likely to underestimate exposure of crop production to drought risks, in particular in areas where water supplies for irrigation are insufficient to fully meet crop water needs in all years due to constraints imposed by hydrology, policies, and/or economics.

@@

Our findings also provide useful guidance for water managers and policymakers seeking to address unsustainable groundwater use in the US High Plains and globally. Consistent with prior theoretical modeling, we demonstrate that the relationship between aquifer thickness and climatic water deficits during the growing season is highly non-linear when accounting for farmers decisions about irrigated production areas. This indicates that sustainable groundwater conservation efforts [@macewan2017hydroecon; @butler2020charting; @elshall2020groundwater] should be targeted in space and time to avoid depletion exceeding critical thresholds which impair farmers' ability to effectively buffer production against drought. Analysis such as ours can provide valuable evidence about these critical thresholds and tipping points to guide proactive management of groundwater stocks, and, in doing so, limit economically damaging transitions from irrigated to rainfed production [@foster2017effects; @deines2020transitions]. For example, several groundwater management agencies in the Texas portion of the HPA have in recent years set out targets to ensure  50% of current aquifer storage and thickness remains in 50 years time (so called 50/50 rule) [@closas2018chronicle]. The effectiveness of such targets could likely be improved substantially by setting conservation targets based on empirical evidence about the level of aquifer thickness that is required to ensure adequate levels of drought protection, both now and with future climate change.
